---
title: "goby_escapevclimb"
author: "Kelly Diamond"
date: "6/2/2021"
output: word_document
---

#Load libraries
```{r load libraries}
library(data.table)
library(ggplot2)
library(emmeans) 
library(lme4)
library(lmerTest)
library(harrellplot)
library(dplyr)
library(geomorph)
library(RRPP)
```

#Create working data.table for escape resposne data
```{r Create working data.table}
fn <- 'escapeData.txt'
file_path <- paste(fn, sep='')
goby <- fread(file_path)

#correct labeling issues for different attack directions
ord <- c('Cranial', 'Lateral', 'Caudal')
goby[, attack_direction:=ifelse(attack_direction=="lateral", "Lateral", attack_direction)]
goby[, attack_direction:=factor(attack_direction, ord)]

#remove juvenile data
goby <- goby[goby$Age!="juvenile", ]

#create species by location column
goby[, Ind:=substr(id, 1,10)]
goby[, Spec_by_loc:=paste0(species, location)]
goby[, Spec_by_reg:=paste0(species, regime)]
goby[, .(N=.N), by=Ind]
goby[, .(N=.N), by=Spec_by_loc]
goby[, .(N=.N), by=Spec_by_reg]
pred_col <- goby[, .(N=.N), by=.(Spec_by_loc, regime, species)]

# change na. action 
make_na_fail <- TRUE
if(make_na_fail==TRUE){
  options(na.action = "na.fail") 
}
```

## Response Frequency
modified from https://github.com/middleprofessor/harrellplot/blob/master/R/fit_model.R
```{r}
#models
fit_resp = glmer(response~climbing_style + attack_direction + (1|species) + (1|location), family="binomial", goby) #full model
fit_resp_attack_only = glmer(response~attack_direction + (1|species) + (1|location), family="binomial", goby) #null for climbing 
fit_resp_climb_only = glmer(response~climbing_style + (1|species) + (1|location), family="binomial", goby) #null for attack
summary(fit_resp)
#X2/p for overall patterns
anova(fit_resp_attack_only,fit_resp)
anova(fit_resp_climb_only,fit_resp)

#create data table of emms
emm.fit_resp <- emmeans(fit_resp, specs=c("climbing_style", "attack_direction"), type="response") 
emm.fit_resp.dt <- data.table(summary(emm.fit_resp))

#compute table of contrasts
emm.resp.climb <- emmeans(fit_resp, specs="climbing_style", type="response") 
emm.resp.attack <- emmeans(fit_resp, specs="attack_direction", type="response") 

ci_diffs.climb <- summary(contrast(
          emm.resp.climb,
          method="pairwise",
          adjust = 'none',
          level=0.95), 
          infer=c(TRUE,TRUE))
ci_diffs.climb.rev <- summary(contrast(
          emm.resp.climb,
          method="revpairwise",
          adjust = 'none',
          level=0.95), 
          infer=c(TRUE,TRUE))
ci_diffs.attack <- summary(contrast(
          emm.resp.attack,
          method="pairwise",
          adjust = 'none',
          level=0.095), 
          infer=c(TRUE,TRUE))
ci_diffs.attack.rev <- summary(contrast(
          emm.resp.attack,
          method="revpairwise",
          adjust = 'none',
          level=0.095), 
          infer=c(TRUE,TRUE))
ci_diffs <- rbind(ci_diffs.climb, ci_diffs.climb.rev, ci_diffs.attack, ci_diffs.attack.rev)
ci_diffs = data.table(ci_diffs)
ci_diffs_resp= ci_diffs[c(3,4,2,7,8,12),]

#my version of the harrell plot
pd <- position_dodge(0.75)
gg_contrasts = ggplot(data=ci_diffs_resp, aes(x=contrast, y=odds.ratio)) +
  geom_hline(yintercept = 1, linetype="dotted", size=1) +
  geom_linerange(aes(ymin = asymp.LCL, ymax = asymp.UCL), color='black', size=1) +
  geom_point(size=3, color='white') +
  geom_point(size=2, color='black') +
  ylab('Ratio') +
  scale_x_discrete("Contrast", breaks = c("none / powerburst","none / inching", "inching / powerburst","Cranial / Lateral", "Cranial / Caudal", "Caudal / Lateral"), labels=c("NC/PB", "NC/IN", "IN/PB", "Cranial/Lateral", "Cranial/Caudal", "Caudal/Lateral")) +
  theme_minimal() + 
  ggpubr::theme_pubclean() + 
  theme(plot.margin = margin(0, 0, 0, 0, "cm"), legend.position = 'bottom') +
  coord_flip()
gg_contrasts

gg_treatments <- ggplot(data=emm.fit_resp.dt, aes(x=climbing_style, y=prob)) +
  geom_errorbar(aes(min=asymp.LCL, max=asymp.UCL, color=attack_direction), 
                 position=pd, width=0.0, size=1.5) +
  geom_point(position=pd, aes(shape=attack_direction), size=5, color='white') +
  geom_point(position=pd, aes(shape=attack_direction), size=3, color='black') +
  ggsci::scale_color_jama(name = "Attack Direction") +
  ggsci::scale_fill_jama() +
  scale_shape(name = "Attack Direction") +
  theme_minimal() +
  scale_y_continuous(name='Response Frequency', labels=scales::percent) +
  scale_x_discrete("Climbing Style", limits = c('inching', 'powerburst', 'none'),
                    labels=c("IN","PB", "NC")) +
  ggpubr::theme_pubclean() +
  theme(plot.margin = margin(0, 0, 0, 0, "cm"), legend.position = 'bottom') +
  coord_flip()
gg_treatments

gg <- cowplot::plot_grid(gg_contrasts, gg_treatments, nrow=2, align = "v")
gg
```

## Performance measures
The following chunk was used to analyze performance measures. For anaylysis of peak velocity and peak acceleration replace 'Escape_Angle_S1' with 'max_vel' or 'max_accel' and change graph labels as appropriate. 
```{r}
#models
fit = lmer(Escape_Angle_S1~climbing_style + attack_direction + (1|species) + (1|location), goby, na.action=na.omit) #full model
fit_attack_only = lmer(Escape_Angle_S1~attack_direction + (1|species) + (1|location), goby, na.action=na.omit) #null for climbing 
fit_climb_only = lmer(Escape_Angle_S1~climbing_style + (1|species) + (1|location), goby, na.action=na.omit) #null for attack

summary(fit)

#X2/p for overall patterns
anova(fit_attack_only,fit)
anova(fit_climb_only,fit)

#create data table of emms
emm.fit <- emmeans(fit, specs=c("climbing_style", "attack_direction"), type="response") 
emm.fit.dt <- data.table(summary(emm.fit))

#compute table of contrasts 
emm.climb <- emmeans(fit, specs="climbing_style", type="response") 
emm.attack <- emmeans(fit, specs="attack_direction", type="response") 

ci_diffs.climb <- summary(contrast(
          emm.climb,
          method="pairwise",
          adjust = 'none',
          level=0.95), 
          infer=c(TRUE,TRUE))
ci_diffs.climb.rev <- summary(contrast(
          emm.climb,
          method="revpairwise",
          adjust = 'none',
          level=0.95), 
          infer=c(TRUE,TRUE))
ci_diffs.attack <- summary(contrast(
          emm.attack,
          method="pairwise",
          adjust = 'none',
          level=0.095), 
          infer=c(TRUE,TRUE))
ci_diffs.attack.rev <- summary(contrast(
          emm.attack,
          method="revpairwise",
          adjust = 'none',
          level=0.095), 
          infer=c(TRUE,TRUE))
ci_diffs <- rbind(ci_diffs.climb, ci_diffs.climb.rev, ci_diffs.attack, ci_diffs.attack.rev)
ci_diffs = data.table(ci_diffs)
ci_diffs= ci_diffs[c(3,4,2,7,8,12),]

#my version of the harrell plot
pd <- position_dodge(0.75)
gg_contrasts = ggplot(data=ci_diffs, aes(x=contrast, y=estimate)) +
  geom_linerange(aes(ymin = lower.CL, ymax = upper.CL), color='black', size=1) +
  geom_hline(yintercept = 1, linetype="dotted", size=1) +
  geom_point(size=3, color='white') +
  geom_point(size=2, color='black') +
  ylab('Effect') +
  scale_x_discrete("Contrast", breaks = c("none - powerburst","none - inching", "inching - powerburst","Cranial - Lateral", "Cranial - Caudal", "Caudal - Lateral"), labels=c("NC-PB", "NC-IN", "IN-PB", "Cranial-Lateral", "Cranial-Caudal", "Caudal-Lateral")) +
  theme_minimal() + 
  ggpubr::theme_pubclean() + 
  theme(plot.margin = margin(0, 0, 0, 0, "cm"), legend.position = 'bottom') +
  coord_flip()
gg_contrasts

gg_treatments <- ggplot(data=emm.fit.dt, aes(x=climbing_style, y=emmean)) +
  geom_errorbar(aes(min=lower.CL, max=upper.CL, color=attack_direction), 
                 position=pd, width=0.0, size=1.5) +
  geom_point(position=pd, aes(shape=attack_direction), size=5, color='white') +
  geom_point(position=pd, aes(shape=attack_direction), size=3, color='black') +
  ggsci::scale_color_jama(name = "Attack Direction") +
  ggsci::scale_fill_jama() +
  scale_shape(name = "Attack Direction") +
  theme_minimal() +
  scale_y_continuous(name='Escape Angle (degrees)') +
  scale_x_discrete("Climbing Style", limits = c('inching', 'powerburst', 'none'),
                    labels=c("IN","PB", "NC")) +
  ggpubr::theme_pubclean() +
  theme(plot.margin = margin(0, 0, 0, 0, "cm"), legend.position = 'bottom') +
  coord_flip()
gg_treatments

gg <- cowplot::plot_grid(gg_contrasts, gg_treatments, nrow=2, align = "v")
gg
```

## Differences among Ss and As localities
```{r localities}
# S. stimpsoni
stimpsoni = goby[goby$species == "Sicyopterus_stimpsoni",]

# Ss.response
summary(aov(response~location, data=stimpsoni))
ss.response = glm(response~location, family="binomial", stimpsoni)
summary(ss.response)
emm.ss.resp <- emmeans(ss.response, specs="location", type="response") 
emm.ss.resp.dt <- data.table(summary(emm.ss.resp))

pd <- position_dodge(0.6)
gg <- ggplot(data=emm.ss.resp.dt, aes(x=location, y=prob)) +
  geom_point(position=pd, aes(color=location), size=7) +
  geom_errorbar(aes(min=asymp.LCL, max=asymp.UCL, color=location), 
                position=pd, width=0.3, size=2) +
  scale_y_continuous(name='Response Frequency', labels=scales::percent) +
  scale_x_discrete("Location", limits = c('nanue1','hakalau','nanue2'), labels=c("BW","AW1", "AW2"))  +
  ggsci::scale_color_jama(name = "Location", breaks=c('nanue1', 'hakalau', 'nanue2'), labels = c('BW = Below Waterfall', 'AW1 = Above Waterfall 1', 'AW2 = Above Waterfall 2')) +
  theme_minimal()+
  theme(axis.text.x=element_text(hjust = 1), plot.margin = margin(0, 0, 0, 1, "cm")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        plot.margin = margin(0.1, 0.1, 0.1, 0.1, "cm"),
        text=element_text(size=20),legend.position=c('none'), legend.text.align = 0, 
        strip.background = element_rect(color = "#cccccc", fill="#cccccc")) +
  theme(axis.line.x = element_line(color="black", size = 1),
        axis.line.y = element_line(color="black", size = 1))+
  NULL
gg

ggsave("ss_response_locality_nopseudo.png", width = 8, height = 8)

#ss.escape angle
summary(aov(Escape_Angle_S1~location, data=stimpsoni, na.action=na.omit))
#ss.velocity
summary(aov(max_vel~location, data=stimpsoni, na.action=na.omit))
#ss.accel
summary(aov(max_accel~location, data=stimpsoni, na.action=na.omit))

# Awaous
awaous = goby[goby$species == "Awaous_stamineus",]

#as.response
summary(aov(response~location, data=awaous))
as.response = glm(response~location, family="binomial", awaous)
emm.as.resp <- emmeans(as.response, specs="location", type="response") 
emm.as.resp.dt <- data.table(summary(emm.as.resp))

pd <- position_dodge(0.6)
gg <- ggplot(data=emm.as.resp.dt, aes(x=location, y=prob)) +
  geom_point(position=pd, aes(color=location), size=7) +
  geom_errorbar(aes(min=asymp.LCL, max=asymp.UCL, color=location), 
                position=pd, width=0.3, size=2) +
  scale_y_continuous(name='Response Frequency', labels=scales::percent) +
  scale_x_discrete("Location", limits = c('waikea', 'hakalau'), labels=c("BW","AW"))  +
  #scale_color_manual(values = c(orange, purple), name=NULL, breaks=c('waikea', 'hakalau'), labels = c('BW = Below Waterfall', 'AW = Above Waterfall')) +
  ggsci::scale_color_jama() +
  theme_minimal()+
  theme(axis.text.x=element_text(hjust = 1), plot.margin = margin(0, 0, 0, 1, "cm")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        plot.margin = margin(0.1, 0.1, 0.1, 0.1, "cm"),
        text=element_text(size=20),legend.position=c('none'), legend.text.align = 0, 
        strip.background = element_rect(color = "#cccccc", fill="#cccccc")) +
  theme(axis.line.x = element_line(color="black", size = 1),
        axis.line.y = element_line(color="black", size = 1))+
  NULL
gg

#as.escape angle
summary(aov(Escape_Angle_S1~location, data=awaous, na.action=na.omit))
#as.velocity
summary(aov(max_vel~location, data=awaous, na.action=na.omit))
#as.accel
summary(aov(max_accel~location, data=awaous, na.action=na.omit))

```

## GM analysis
# Setup
```{r libraries}
library(ape)
library(MASS)
library(shapes)
library(rgl)
library(geomorph)
library(RRPP)
library(scatterplot3d)
library(vegan)
library(ggplot2)
library(Hmisc)
library(lme4)
```

# Running size corrected gpa for body shape landmarks
```{r}
#upload landmark data and save as array for gpagen fxn
LMdata<-read.csv('lm_data.csv', header = TRUE)
TPS = as.matrix(LMdata[,2:25]) 
TPS = arrayspecs(TPS, 12, 2) 

#procrustes superimposition
y.gpaTPS = gpagen(TPS, PrinAxes = TRUE, max.iter = NULL, ProcD = TRUE, Proj = TRUE)
plot(y.gpaTPS)

#size correction
size<-read.csv("sizes.csv")
GDF <- geomorph.data.frame(shape = y.gpaTPS$coords, CS = size$SL, logCS = log(size$SL))
sizecorrection =procD.lm(shape ~ CS, f2 = NULL, f3=NULL, 
logsz = TRUE, data=GDF, iter=149, print.progress = FALSE)
plot(sizecorrection, method="PredLine")   

#procrustes impostion for size corrected data
morph<-procD.lm(shape ~ CS, data=GDF,iter=1000, RRPP = TRUE)
summary(morph)#gives the p value associated with size in this data set
shape.resid<-arrayspecs(morph$residuals,p=dim(y.gpaTPS$coords)[1],k=dim(y.gpaTPS$coords)[2])
adj.shape<-shape.resid+array(y.gpaTPS$consensus,dim(shape.resid))

#adding groups to dataset
gp <- interaction(LMdata$species)
col.gp <- c('#374E55FF', '#DF8F44FF', '00A1D5F')
names(col.gp) <- levels(gp)
col.gp <- col.gp[match(gp, names(col.gp))] # col.gp must NOT be a factor

#finally, we get what we want: the allometry free data (PC1 vs PC2)
corrected.pca_L<-plotTangentSpace(adj.shape, groups= col.gp, axis1=1, axis2=2, label=T, legend = FALSE) #flip axes to get pics for figures for ms then pull out in ps
corrected.pca<-plotTangentSpace(adj.shape, groups= col.gp, axis1=1, axis2=2) 
corrected.pca$pc.summary #PCs with % explained

#export points
corrected.pca$pc.shapes
corrected.pca$rotation

```

#stats on PC scores
```{r PC stats}
#upload pc scores with lables for species/ location/ predator regime
goby_pcs <- read.csv("pc_scores_labeled.csv", header = TRUE)

#Does PC1 differ by climbing sytle
fit.pc1 <- lmer(PC1 ~ climbing_style + (1|species) + (1|location), data = goby_pcs)
summary(fit.pc1)
summary(aov(PC1 ~ climbing_style, data = goby_pcs))

#Does PC2 differ by climbing style
fit.pc2 <- lmer(PC2 ~ climbing_style + (1|species) + (1|location), data = goby_pcs)
summary(fit.pc2)
summary(aov(PC2 ~ climbing_style, data = goby_pcs))

#Does PC3 differ by climbing style
fit.pc3 <- lmer(PC3 ~ climbing_style + (1|species) + (1|location), data = goby_pcs)
summary(fit.pc3)
summary(aov(PC3 ~ climbing_style, data = goby_pcs))

#Does PC4 differ by climbing style
fit.pc4 <- lmer(PC4 ~ climbing_style + (1|species) + (1|location), data = goby_pcs)
summary(fit.pc4)

```

#Tukey for climbing style (for supplemental only)
```{r}
tpc1 <- aov(PC1~ climbing_style, goby_pcs)
TukeyHSD(tpc1, "climbing_style")
summary(tpc1)

tpc2 <- aov(PC2~ climbing_style, goby_pcs)
TukeyHSD(tpc2, "climbing_style")
summary(tpc2)

tpc3 <- aov(PC3~ climbing_style, goby_pcs)
TukeyHSD(tpc3, "climbing_style")
summary(tpc3)

```

#pretty plots
```{r pretty plots}
#make new df with pcscores
goby_pcs <- read.csv("pc_scores_labeled.csv", header = TRUE)

# +/_ axis warps
pdf("pc1v2warps.pdf")
corrected.pca<-plotTangentSpace(adj.shape, groups= col.gp, axis1=1, axis2=2) 
dev.off()

pdf("pc2v1warps.pdf")
corrected.pca<-plotTangentSpace(adj.shape, groups= col.gp, axis1=2, axis2=1) 
dev.off()

pdf("pc3v1warps.pdf")
corrected.pca<-plotTangentSpace(adj.shape, groups= col.gp, axis1=3, axis2=1) 
dev.off()

#color scheme
orange <- rgb(230, 97, 1, max = 255) #inching
purple <- rgb(94, 60, 153, max = 255) #none
grey <- rgb(150, 150, 150, max = 255) #powerburst
c_sch <- c(orange, grey, purple)

#PC1 vs PC2
pd <- 0.5
gg <- ggplot(data=goby_pcs, aes(x=PC1, y=PC2, color=climbing_style)) +
  geom_point(size=5) +
  xlab("PC1 (48.2%)")+
  ylab("PC2 (14.0%)")+
  theme_minimal() +
  scale_color_manual(values = c_sch, name=NULL, breaks=c('none', 'powerburst', 'inching'), 
                     labels = c('Non-climbing', 'Powerburst', 'Inching')) +
  theme(axis.text.x=element_text(hjust = 1),
        plot.margin = margin(0, 0, 0, 1, "cm")) +
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(),
        panel.spacing = unit(2, "lines"),
        plot.margin = margin(0.1, 0.1, 0.1, 0.1, "cm"), text=element_text(size=20),legend.position=c('top'), legend.text.align = 0, 
        strip.background = element_rect(color = "#cccccc", fill="#cccccc")) +
  theme(axis.line.x = element_line(color="black", size = 1),
        axis.line.y = element_line(color="black", size = 1))+
  NULL
gg


#PC1 vs PC3
pd <- 0.5
gg <- ggplot(data=goby_pcs, aes(x=PC1, y=PC3, color = climbing_style, shape = species)) +
  geom_point(size=5) +
  xlab("PC1 (48.2%)")+
  ylab("PC3 (13.0%)")+
  theme_minimal() +
  scale_color_manual(values = c_sch, name="Climbing Style", breaks=c('none', 'powerburst', 'inching'), 
                     labels = c('Non-climbing', 'Powerburst', 'Inching')) +
  scale_shape_manual(values = c(19,17,15,6,7,8), name='Species', breaks=c("Stenogobius hawaiiensis","Cotylopus acutipinnis", "Lentipes concolor", "Awaous stamineus", "Sicyopterus lagocephalus", "Sicyopterus stimpsoni"), labels=expression(paste(italic("S. hawaiiensis")), paste(italic("C. acutipinnis")), paste(italic("L. concolor")), paste(italic("A. stamineus")), paste(italic("S. lagocephalus")), paste(italic("S. stimpsoni")))) +
  theme(axis.text.x=element_text(hjust = 1),
        plot.margin = margin(0, 0, 0, 1, "cm")) +
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        panel.background = element_blank(),
        panel.spacing = unit(2, "lines"),
        plot.margin = margin(0.1, 0.1, 0.1, 0.1, "cm"), text=element_text(size=20),legend.position=c('right'), legend.text.align = 0, 
        strip.background = element_rect(color = "#cccccc", fill="#cccccc")) +
  theme(axis.line.x = element_line(color="black", size = 1),
        axis.line.y = element_line(color="black", size = 1))+
  NULL
gg

```


